### 2013-07-01 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
updating parallel algorithm explanation

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. Their
algorithm is presented in further detail in the reference and was
effectively implemented verbatim for this work. In their work they
identify two data sets that are required to be communicated: the
sharing of particles that are transported from one domain to another
and therefore from one processor to another and a global communication
that signals if particle transport has been completed on all
processors. Both of these processes will be addressed at a high level
along with how the Monte Carlo data structures are
constructed in parallel.

\subsection{Parallel Transport Domain and Source }
\label{subsec:domain_generation}
To utilize the parallel transport algorithm, we must generate the
parallel transport domain. For the Neumann-Ulam method, the domain
consists of all states in the system that are local, and the
probabilities and weights for all state transitions possible in the
local domain. At the core of this representation is the Neumann-Ulam
decomposition of the linear operator. Given an input decomposition for
the linear operator, by definition the Neumann-Ulam decomposition will
have the same parallel decomposition. In addition, any modification
made to the linear operator through preconditioning or relaxation
parameters will also modify the Neumann-Ulam decomposition. All
possible transitions are described by the local graph of the sparse
input matrix. In addition, any left preconditioning or relaxation
parameters will modify the right hand side of the linear system and
therefore the fixed source in the Monte Carlo calculation. This source
vector will have the same parallel decomposition as the input matrix
and therefore all of the birth states for the histories will exist for
at least a single transition event within the local domain.

@@

\begin{algorithm}[h!]
  \caption{Parallel Neumann-Ulam Algorithm}
  \label{alg:parallel_mc_algorithm}
  \begin{algorithmic}[1]
    \State get list of neighbor processors 
    \Comment{Each neighbor owns an adjacent subdomain}
    \ForAll{neighbors}
    \State post nonblocking receive for maximum history buffer size
    \State allocate history buffer
    \EndFor
    \State \textit{historiesCompleted} = 0
    \Comment{local+child finished histories}
    \State \textit{localProcessed} = 0
    \Comment{local trajectories computed (not necessarily finished)}
    \State calculate parent and children processor ID numbers in
    binary tree
    \ForAll{child processes}
    \State post nonblocking receive for \textit{historiesCompleted} tally
    \EndFor
    \State post nonblocking receive for stop message from parent
    \While{stop flag not set}
    \If{any local histories in source or stack}
    \State \textbf{LocalHistoryTransport()}
    \State ++\textit{localProcessed}
    \EndIf
    \If{checkPeriod == \textit{localProcessed} || no local histories}
    \State \textbf{ProcessMessages()}
    \State \textit{localProcessed} = 0
    \EndIf
    \If{no local histories}
    \State \textbf{ControlTermination()}
    \EndIf
    \EndWhile
    \State cancel outstanding nonblocking requests
    \State free all history buffers
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Full-Clip Neumann-Ulam}
\label{subsec:full_clip}

>>>>

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. For
considerably more detail, the algorithm presented can be found in
\citep{brunner_efficient_2009} and was effectively implemented
verbatim for this work. In their work they identify two data sets that
are required to be communicated: the sharing of particles that are
transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Both of these
processes will be addressed at a high level along with how the Monte
Carlo data structures they require are constructed in parallel.

\subsection{Parallel Transport Domain and Source }
\label{subsec:domain_generation}
To utilize a parallel transport algorithm, we must generate the
required data with the correct parallel decomposition. For the
Neumann-Ulam method, the transport domain consists of all states in
the system that are local, and the probabilities and weights for all
state transitions possible in the local domain. At the core of this
representation is the Neumann-Ulam decomposition of the linear
operator. Given an input decomposition for the linear operator, by
definition the Neumann-Ulam decomposition will have the same parallel
decomposition. In addition, any modification made to the linear
operator through preconditioning or relaxation parameters will also
modify the Neumann-Ulam decomposition. All possible transitions are
described by the local graph of the sparse input matrix. In addition,
any left preconditioning or relaxation parameters will modify the
right hand side of the linear system and therefore the fixed source in
the Monte Carlo calculation. This source vector will have the same
parallel decomposition as the input matrix and therefore all of the
birth states for the histories will exist for at least a single
transition event within the local domain.

Compared to the serial construction of the Neumann-Ulam decomposition,
data for all states that is required to be on process must be
collected in parallel. For the pure domain decomposition case, given
an input parallel decomposition for the Neumann-Ulam decomposition,
for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
adjacency graph of a single state in the neutron diffusion matrix for
the model problem presented in Appendix~\ref{chap:diffusion_problem}.
\begin{figure}[htpb!]
  \begin{center}
    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
  \end{center}
  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
    \textit{Adjacent boundary states are collected by traversing the
      graph one step for all local boundary states and collecting the
      equations for those states that are not local. For a given
      transition in the random walk for the diffusion problem, a
      history at mesh point $(i,j)$ may transition to all adjacent
      mesh points including itself, corresponding to a global state
      transition from $m$ to $n$ in the linear system.}}
  \label{fig:diffusion_graph}
\end{figure}

In this example, we currently reside in state $m$ of the system,
directly correlating to physical location $(i,j)$ in the mesh (the
center node). The discretization stencil of the diffusion problem
dictates the structure of this graph and the states to which a history
may transition. If we play the Monte Carlo game to move to a new state
$n$, then we may move to any of the nodes in this graph, including the
node at which we started. If grid point $(i,j)$ in this example is on
the right of the local domain boundary and we transition to the right
to state $n$ corresponding to mesh point $(i+1,j)$, we are now in a
state that is owned by the adjacent domain. We must then gather the
data in the Neumann-Ulam decomposition from the neighboring process
that owns grid point $(i+1,j)$. Doing this provides us with the data
required by the estimators permitting $P_{mn}$ and $W_{mn}$ to be
computed locally for this particular boundary transition. For the
parallel source, these adjacent states are not required as histories
may only be born in states in the local domain. Once the local
Neumann-Ulam decomposition has been generated with the proper data
collected from adjacent domains, Monte Carlo transport may proceed in
parallel.

@@

Here we present in detail Brunner and Brantley's 2009 algorithm in
detail and discuss how it is adapted to parallelize the Neumann Ulam
method. Presented in Algorithm~\ref{alg:parallel_mc_algorithm}, the
top level sequence performs the task of managing history transport
through the local domain, communication of histories to adjacent
domains, and the completion of transport. For each of these tasks,
additional algorithms, shown in bold in
Algorithm~\ref{alg:parallel_mc_algorithm}
(e.g. \textbf{LocalHistoryTransport}), are presented for additional
detail.

\begin{algorithm}[h!]
  \caption{Parallel Neumann-Ulam Algorithm}
  \label{alg:parallel_mc_algorithm}
  \begin{algorithmic}[1]
    \State get list of neighbor processors 
    \Comment{Each neighbor owns an adjacent subdomain}
    \ForAll{neighbors}
    \State post nonblocking receive for maximum history buffer size
    \State allocate history buffer
    \EndFor
    \State \textit{historiesCompleted} = 0
    \Comment{local+child finished histories}
    \State \textit{localProcessed} = 0
    \Comment{local trajectories computed (not necessarily finished)}
    \State calculate parent and children processor ID numbers in
    binary tree
    \ForAll{child processes}
    \State post nonblocking receive for \textit{historiesCompleted} tally
    \EndFor
    \State post nonblocking receive for stop message from parent
    \While{stop flag not set}
    \If{any local histories in source or stack}
    \State \textbf{LocalHistoryTransport()}
    \State ++\textit{localProcessed}
    \EndIf
    \If{message check period == \textit{localProcessed} || no local histories}
    \State \textbf{ProcessMessages()}
    \State \textit{localProcessed} = 0
    \EndIf
    \If{no local histories}
    \State \textbf{ControlTermination()}
    \EndIf
    \EndWhile
    \State cancel outstanding nonblocking requests
    \State free all history buffers
  \end{algorithmic}
\end{algorithm}

@@

\subsection{Subdomain Neumann-Ulam}
\label{subsec:full_clip}

Full-clip actually uses the solver as a true subdomain solver, just
like you would in a Schwarz method. In fact, this is a stochastic
realization of a Richardson iteration - they are very similar in those
regards.

Overlap is not effective because histories apparently dump most of
their useful information within the first few steps (which is why we
can use such a large weight cutoff and there is an insensitivity to
weight cutoff). So keeping them around to do extra work is not
improving iterative performance - thus increasing compute times. This
is true even for ill-conditioned cases where adding overlap might keep
a number of histories within the domain for significantly longer.

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_strong_subdomain.pdf}
  \end{center}
  \caption{\textbf{Strong scaling absolute efficiency for pure domain
      decomposition.} \textit{The subdomain Monte Carlo solver
      improves the parallel efficiency and therefore time to solution
      by eliminating all parallel communication during the Monte Carlo
      solve.}}
  \label{fig:titan_strong_subdomain}
\end{figure}

\begin{figure}[htpb!]
  \begin{center}
    \includegraphics[width=6in]{chapters/parallel_mc/titan_weak_subdomain.pdf}
  \end{center}
  \caption{\textbf{Weak scaling absolute efficiency for pure domain
      decomposition.} \textit{The subdomain Monte Carlo solver
      improves the parallel efficiency and therefore time to solution
      by eliminating all parallel communication during the Monte Carlo
      solve.}}
  \label{fig:titan_weak_subdomain}
\end{figure}
